Developing a model for loan risk estimation by Emadeldin_Fawaz
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading Needed Packages
library(readr)
library(ggplot2) 
library(gridExtra) 
library(plyr)
library(ggplot2)
library(ggcorrplot)
library(RColorBrewer)
library(plotrix)
library(dplyr)
library(memisc)
library(forcats)
library(grid)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data

df <- read_csv("prosperLoanData.csv")

```
##


# Introduction

This project was conducted as part of a nano degree at Udacity. Because data 
visualization and data science is a passion, I decided to limit the time to 
invest in this project to 30 working hours. 
The data used in this project is called "Loan Data Form Prosper" and it was 
supplied with the project instructions. However, it is also available online 
(please see the references).
In this project, I aimed at visualizing the different variable separately and
against each other. The idea was to try to identify a pattern or a 
relationship between the different variables. I aimed to develop and test a
model that helps to predict one of the important variables (from my point of 
view). Eventually, I was successful in developing a simple 5 variable model 
that predicts the yield on loans quite successfully. 


# Preliminary data investigation and understanding the dataframe


```{r echo=FALSE, Preliminary_data_investigation_one}
str(df)
summary(df)

```

The data is relatively large (113937 observations of 81 variables). 
I am concerned that running any code on such huge dataframe will take too long. 
I will generate a sub-dataframe of randomly selected 12000 observations 
(around 10% of the total observations) to be my working dataframe 
(named: subfd). 
I will explore the subset of the dataframe to see if I can come up with 
a plausible hypothesis. Then I will test my hypothesis on the full dataframe.



```{r echo=FALSE, Preliminary_data_investigation_two}

subdf <- df[sample(1:nrow(df), 12000),]

summary(subdf)

colnames(subdf)

```


I compared some indicators between the full df and the subdf. Ignoring possible
outliers, it looks like that my sub-dataframe could represent the original 
dataframe. Among the numerical indicators I looked at:
Term, BorrowerRate, ProsperRating (numeric), EmploymentStatusDuration, AvailableBankcardCredit, StatedMonthlyIncome, and LoanOriginalAmount.



Interesting to notice that some people with a monthly income of more than 
$250,000 are applying for loans with max $35000.


I will focus on some variables that I think they are of particular interest to 
the model I am trying to build:-

Some variables are reflecting each other (like ProsperRating (numeric), 
ProsperRating (Alpha) and ProsperScore). So I will choose one of them in such 
cases. Some variables are included in others (like LenderYield: The Lender 
yield is equal to the interest rate on the loan less the servicing fee). 
So I will use only the LenderYield and not the BorrowerRate. 
GroupKey includes if the borrower was in a group or not. 
Thus the variable CurrentlyInGroup could be omitted.



Variables that could be of particular interest to my model are:



A.	Numerical

1.	"Term"                                
2.	"LenderYield"                        
3.	"EstimatedReturn"  
4.	"ProsperScore"                       
5.	"EmploymentStatusDuration"            
6.	"CreditScoreRangeLower"              
7.	"CreditScoreRangeUpper"               
8.	"CurrentCreditLines"                  
9.	"OpenCreditLines"                    
10.	"OpenRevolvingAccounts"              
11.	"OpenRevolvingMonthlyPayment"         
12.	"InquiriesLast6Months"               
13.	"TotalInquiries" 
14.	"AmountDelinquent"                    
15.	"DelinquenciesLast7Years"            
16.	"PublicRecordsLast10Years"            
17.	"PublicRecordsLast12Months"          
18.	"RevolvingCreditBalance"              
19.	"BankcardUtilization"                
20.	"AvailableBankcardCredit"             
21.	"TotalTrades"                        
22.	"TradesNeverDelinquent (percentage)"  
23.	"TradesOpenedLast6Months"            
24.	"DebtToIncomeRatio"                   
25.	"IncomeRange"                        
26.	"StatedMonthlyIncome"                
27.	"TotalProsperLoans"                  
28.	"OnTimeProsperPayments"              
29.	"ProsperPaymentsLessThanOneMonthLate" 
30.	"ProsperPaymentsOneMonthPlusLate"    
31.	"ProsperPrincipalBorrowed"            
32.	"ProsperPrincipalOutstanding"        
33.	"ScorexChangeAtTimeOfListing"         
34.	"LoanCurrentDaysDelinquent"          
35.	"LoanMonthsSinceOrigination"         
36.	"LoanOriginalAmount"                 
37.	"MonthlyLoanPayment"                 
38.	"LP_CustomerPayments"                 
39.	"LP_CustomerPrincipalPayments"       
40.	"LP_InterestandFees"                 
41.	"LP_NetPrincipalLoss"                
42.	"PercentFunded"                       
43.	"Recommendations"                    
44.	"InvestmentFromFriendsCount"          
45.	"InvestmentFromFriendsAmount"        
46.	"Investors


	
B.	Categorical

1.	"CreditGrade"  
2.	"LoanStatus"
3.	"ListingCategory (numeric)"           
4.	"Occupation"                          
5.	"EmploymentStatus"                   
6.	"IsBorrowerHomeowner"                
7.	"IncomeVerifiable"                
8.	"LoanOriginationQuarter"
9.	"EmploymentStatus"



ListingCategory
The category of the listing that the borrower selected when posting their 
listing: 0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 
3 - Business, 4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 
8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 
12 - Green Loans, 13 - Household Expenses, 14 - Large Purchases, 
15 - Medical/Dental, 16 - Motorcycle, 
17 - RV, 18 - Taxes, 19 - Vacation, 20 - Wedding Loans



I will use box plots to explore the numerical variables and histograms to explore
the categorical variables.



# Univariate Plots Section

I will use boxplots for plotting the numerical variables and 
histograms for plotting the categorical variables.
 

```{r echo=FALSE, Univariate_Plots_one}
# Ploting Numerical variables

NumPlot1 = qplot(y= InquiriesLast6Months, data = subdf, geom="boxplot")
NumPlot2 = qplot(y= TotalInquiries, data = subdf, geom="boxplot")         

grid.arrange(NumPlot1, NumPlot2, ncol = 2)

with (subdf, cor.test(InquiriesLast6Months, TotalInquiries))

```

As expected, there is a very high correlation (rho > 0.7) between 
InquiriesLast6Months and TotalInquiries. Accordingly, in the model I will only 
use the most recent variable: InquiriesLast6Months



```{r echo=FALSE, Univariate_Plots_two}
# Ploting Numerical variables

NumPlot3 = qplot(y= AvailableBankcardCredit, data = subdf, geom="boxplot")             
NumPlot4 = qplot(y= DebtToIncomeRatio, data = subdf, geom="boxplot")           

grid.arrange(NumPlot3, NumPlot4, ncol = 2)


summary(subdf$AvailableBankcardCredit)


summary(subdf$DebtToIncomeRatio)

```

The majority of the people applying for the loans have very low dept to income 
ratio (75% are below 0.32) and very low Available Bank Card Credit 
(75% below $13,245). This is expected because the maximum loan in the data has
the value of $35,000. With such small loans it is not expected that the 
majority of the customers are high earners.



```{r echo=FALSE, Univariate_Plots_three}
# Ploting Numerical variables


NumPlot5 = qplot(y= StatedMonthlyIncome, data = subdf, geom="boxplot")            
NumPlot6 = qplot(y= LoanOriginalAmount, data = subdf, geom="boxplot")                     
NumPlot7 = qplot(y= MonthlyLoanPayment, data = subdf, geom="boxplot")
NumPlot8 = qplot(y= EmploymentStatusDuration, data = subdf, geom="boxplot")

grid.arrange(NumPlot5, NumPlot6, NumPlot7, NumPlot8, ncol = 2)

summary(subdf$StatedMonthlyIncome)
summary(subdf$LoanOriginalAmount)

```

75% of the loan applicants have monthly income below $6874. I believe the 
majority of the loan applicants are employees or small scale entrepreneurs.
Too much outliers are noted in these variables.
                  

```{r echo=FALSE, Univariate_Plots_four}
# Ploting categorical variables

CatPlot1 = qplot(x= CreditGrade, data = subdf)
CatPlot2 = qplot(x= subdf$`ListingCategory (numeric)`, binwidth = 1)

grid.arrange(CatPlot1, CatPlot2, ncol = 2)


```



```{r echo=FALSE, Univariate_Plots_five}

CatPlot3 = qplot(x= subdf$Occupation)
CatPlot4 = qplot(x= EmploymentStatus, data = subdf)

grid.arrange(CatPlot3, CatPlot4, ncol = 1)


```

```

count(subdf$Occupation)
count(subdf$EmploymentStatus)

```

```{r echo=FALSE, Univariate_Plots_six}

CatPlot6 = qplot(x= IsBorrowerHomeowner, data = subdf)
CatPlot7 = qplot(x= IncomeVerifiable, data = subdf)


grid.arrange(CatPlot6, CatPlot7, ncol = 2)


```






library(plotrix)

pie3D(as.data.frame(count(subdf$Occupation))$freq, labels = as.data.frame(count(subdf$Occupation))$x,  main = "Occupations", explode=0.1, 
      radius=2, labelcex = 0.5,  start=0.7)



Too many occupations. The pie chart will not be adequate to summarize 
all data.





```{r echo=FALSE, Univariate_Plots_seven}

library(forcats)


ggplot(subdf, aes(x = fct_infreq(Occupation), fill = fct_infreq(Occupation))) +
  geom_bar(width = 1, colour = "black", show.legend = FALSE) +
  xlab("Customer Occubations") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))



```


## Univariate Analysis

I used the univariate analysis to understand the characteristics of Loan 
Borrowers.

The majority of the customers have low dept to income 
ratio (75% are below 0.32) and low Available Bank Card Credit 
(75% below $13,245). 
This is expected because the maximum loan in the data has the value of $35,000.
With such small loans it is not expected that the majority of the customers 
are high earners.

75% of the loan applicants have monthly income below $6874. 
I believe the majority of the loan applicants are employees or small scale 
entrepreneurs.

Dept consolidation was the most frequent reason for applying for a loan. 
Business and home improvement were the 2nd and 3rd frequent loan reasons 
(respectively) in the data. Loan grades were not applicable for most of the
loan borrowers (8943 out of 12000 in the subseted dataframe). However, most 
of the known grades were of the middle range (C-D).

As expected, most of the borrowers are employees. Their income could be 
verified. Owning a house or not was almost equally distributed among applicants.



# Bivariate Plots Section

The most important dependent variables are:
  CreditGrade
  LoanStatus
  LenderYield
  
Other variables like estimated return will be correlated to one of these 
variables (by definition).

```{r echo=FALSE, Bivariate_Plots_one}
#understanding the vriables

with(subdf, cor.test(LenderYield, EstimatedReturn))


```
For instance, the Correlation coefficient between the Estimated return and 
Lender yield is 0.815 which is significantly high.


The LoanStatus & cridetGrade variable are categorical variables. It is better
to transform them to numerical variables before proceeding


#understanding the vriables

count(subdf$CreditGrade)

count(subdf$LoanStatus)

```{r echo=FALSE, Bivariate_Plots_two}

subdf$LoanStatusNum <- 
  ifelse(subdf$LoanStatus == 'Chargedoff', 1, 
    ifelse(subdf$LoanStatus == 'Defaulted', 2,
      ifelse(subdf$LoanStatus == 'Past Due (>120 days)', 3,
        ifelse(subdf$LoanStatus == 'Past Due (91-120 days)', 3,
          ifelse(subdf$LoanStatus == 'Past Due (61-90 days)', 3,
            ifelse(subdf$LoanStatus == 'Past Due (31-60 days)', 4,
              ifelse(subdf$LoanStatus== 'Past Due (16-30 days)', 5,
                ifelse(subdf$LoanStatus == 'Past Due (1-15 days', 6,
                 ifelse(subdf$LoanStatus == 'Current', 7,       
                  ifelse(subdf$LoanStatus == 'FinalPaymentInProgress', 8,
                    ifelse(subdf$LoanStatus == 'Completed', 9,
                      NA)))))))))))


```


```{r echo=FALSE, Bivariate_Plots_three}
subdf$CreditGradeNum <- ifelse(subdf$CreditGrade == 'NC', 1, 
                               ifelse(subdf$CreditGrade== 'HR', 2,
                                ifelse(subdf$CreditGrade== 'E', 3,
                                 ifelse(subdf$CreditGrade== 'D', 4,
                                  ifelse(subdf$CreditGrade== 'C', 5,
                                   ifelse(subdf$CreditGrade== 'B', 6,
                                    ifelse(subdf$CreditGrade== 'A', 7,
                                     ifelse(subdf$CreditGrade== 'AA', 8,       
                                            NA))))))))

```




## Finding most important independent variables

count(subdf$Recommendations)
 
count(subdf$PublicRecordsLast12Months)

count(subdf$InquiriesLast6Months)

count(subdf$TotalProsperLoans)

summary(subdf$OnTimeProsperPayments)

summary(subdf$PercentFunded)

summary(subdf$Term)


 The majority of the applicants do not have recommendations and public records.
 Total prosper loans is mainly 1 or 2 (expected for employees) and most of 
 its information is not available. The loans are mainly fully funded by the 
 lender.
 

```{r echo=FALSE, Bivariate_Plots_four}



p1 = qplot(x = as.character(CreditGradeNum), 
      y =  EmploymentStatusDuration, 
      data = subdf, 
      geom = 'boxplot') +
      ylim(0, 500)

p2 = qplot(x = as.character(CreditGradeNum), 
      y =  CurrentCreditLines, 
      data = subdf, 
      geom = 'boxplot') +
      ylim(0,30)

p3= ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                        y = DebtToIncomeRatio)) + 
        geom_boxplot() + 
        ylim(0, 1.5)

p4 = ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = StatedMonthlyIncome)) +
        geom_boxplot() +
        ylim(0,15000)

p5=   ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = InquiriesLast6Months)) +
        geom_boxplot() +
        ylim(0,10)

p6 = ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = AmountDelinquent)) +
        geom_boxplot() +
        ylim(0,50000)    


p7 = ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = AvailableBankcardCredit)) +
        geom_boxplot() +
        ylim(0,10000)  


p8 = ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = OnTimeProsperPayments)) +
        geom_boxplot()+
        ylim(0,30)   

p9 = ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                         y = LenderYield)) +
        geom_boxplot()   


grid.arrange(p1, p2, p3, p4, p5, p6,p7, p8, p9, ncol = 2)

```

It looks like only two credit grades are strongly influenced by the duration 
of employment. AA credit grade has higher employment status duration and 
grade HR has lower employment status duration.The rest of the grades are
not showing a particular pattern.

It is more obvious that the increase in current credit lines is associated 
 with higher credit grades.

Dept to income ratio is not playing a significant role in the credit grade.
This might be expected in this case because the data is more about employees.
Dept to income ratio might be more decisive in corporate loans. It is hard to 
verify Dept in the case of people (unless they volunteeringly reveal) .

Stated monthly income is playing a role in deciding the credit grade. To some 
extent, the boxes are moving up with the better credit grades.

It looks like the increase in Inquiries in the Last 6 Months is, to some extent, 
is associated with increase in the credit grade.

No risk no fun. As the credit grade decreases the lender yield increases


```{r echo=FALSE, Bivariate_Plots_five}



p1 = qplot(x = as.character(LoanStatusNum), 
      y =  Term, 
      data = subdf, 
      geom = 'boxplot') +
  ylim(0, 60)

p2 = qplot(x = as.character(LoanStatusNum), 
      y =  EmploymentStatusDuration, 
      data = subdf, 
      geom = 'boxplot') +
  ylim(0,30)

p3= ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                        y = CurrentCreditLines)) + 
        geom_boxplot() + 
        ylim(0, 30)
p4 = ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                         y = InquiriesLast6Months)) +
        geom_boxplot() + 
        ylim(0, 15)

p5=   ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                         y = StatedMonthlyIncome)) +
        geom_boxplot() + 
        ylim(0, 20000)

p6 = ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                         y = OnTimeProsperPayments)) +
        geom_boxplot() + 
        ylim(0, 80)   

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)


```
  
The majority of the loans were for 36 months. Loan Term is not affecting the 
status.

Duration of employment status is not affecting the loan status.

Loan status is getting better with increase in Current credit line.

Inquiries in the last 6 months do not affect loan status.

Stated monthly income is, to some extent, a decisive factor that enhances
loan status.

On time prosper payment is not showing a specific pattern in relation to
loan status.


```{r echo=FALSE, Bivariate_Plots_six}

p1 = qplot(x = IsBorrowerHomeowner, 
      y =  LoanStatusNum, 
      data = subdf, 
      geom = 'boxplot') 

p2 = qplot(x = IncomeVerifiable, 
      y =  LoanStatusNum, 
      data = subdf, 
      geom = 'boxplot')


grid.arrange(p1, p2, ncol = 1)


```
 
 Home ownership and income verifiability are not significantly impacting 
 loan status.
 

```{r echo=FALSE, Bivariate_Plots_seven}



p1= ggplot(subdf, aes(x = LenderYield, 
                        y = LoanOriginalAmount)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p2= ggplot(subdf, aes(x = LenderYield, 
                        y = EmploymentStatusDuration)) + 
        geom_point(alpha = 1/50) + 
        ylim(0, 30) +
       geom_smooth(method="lm")


p3= ggplot(subdf, aes(x = LenderYield, 
                        y = CurrentCreditLines)) + 
        geom_point(alpha = 1/50) + 
        ylim(0, 30)+
        geom_smooth(method="lm")

p4 = ggplot(subdf, aes(x = LenderYield, 
                         y = InquiriesLast6Months)) +
        geom_point(alpha = 1/50) + 
        ylim(0, 15) + 
        geom_smooth(method="lm")

p5=   ggplot(subdf, aes(x = LenderYield, 
                         y = StatedMonthlyIncome)) +
        geom_point(alpha = 1/50) + 
        ylim(0, 20000) + 
        geom_smooth(method="lm")

p6 = ggplot(subdf, aes(x = LenderYield, 
                         y = OnTimeProsperPayments)) +
        geom_point(alpha = 1/50) + 
        ylim(0, 80) +
        geom_smooth(method="lm")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)

  

```

Lender yield is increasing with Inquiries last 6 months and ontime prosper 
payment. 
Lender yield is decreasing with loan original amount, Employment status 
duration, current credit lines and stated monthly income. 

loan original amount and stated monthly income showed the strongest slope 
indicating that they have a stronger effect on lender yield.


```{r echo=FALSE, Bivariate_Plots_eight}


p1= ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                        y = AmountDelinquent)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p2= ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                        y = AvailableBankcardCredit)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")


p3= ggplot(subdf, aes(x = as.character(CreditGradeNum), 
                        y = MonthlyLoanPayment)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p4= ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                        y = AmountDelinquent)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p5= ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                        y = AvailableBankcardCredit)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")


p6= ggplot(subdf, aes(x = as.character(LoanStatusNum), 
                        y = MonthlyLoanPayment)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p7= ggplot(subdf, aes(x = LenderYield, 
                        y = AmountDelinquent)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")

p8= ggplot(subdf, aes(x = LenderYield, 
                        y = AvailableBankcardCredit)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")


p9= ggplot(subdf, aes(x = LenderYield, 
                        y = MonthlyLoanPayment)) + 
        geom_point(alpha = 1/50) + 
        geom_smooth(method="lm")



grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)



```

These three variables: AmountDelinquent, AvailableBankcardCredit and 
MonthlyLoanPayment are impacting the important dependent variables.




## Bivariate Analysis

Three variables are potential important outcomes in the model I am trying to
develop. These are:

CreditGrade
LoanStatus
LenderYield

Out of the tested dependent variables, only 8 were identified that might 
influence the targeted dependent variables. These are:
LoanOriginalAmount
CurrentCreditLines
InquiriesLast6Months
StatedMonthlyIncome
OnTimeProsperPayments
AmountDelinquent
AvailableBankcardCredit
MonthlyLoanPayment

In the next section I will try to plot the most plausible relationships 
and develop and test a model.





# Multivariate Plots Section

## Extracting data for the model 

```{r echo=FALSE, Multivariate_Plots_one}

Model <-
  data.frame(subdf$CreditGradeNum,
         subdf$LoanStatusNum,
         subdf$LenderYield,
         subdf$LoanOriginalAmount,
         subdf$CurrentCreditLines,
         subdf$InquiriesLast6Months,
         subdf$StatedMonthlyIncome,
         subdf$OnTimeProsperPayments,
         subdf$AmountDelinquent,
         subdf$AvailableBankcardCredit,
         subdf$MonthlyLoanPayment)


```

## Plotting correlations Matrix

```{r echo=FALSE, Multivariate_Plots_two}

# Correlation matrix

# Install package
if(!require(devtools)) install.packages("devtools")
devtools::install_github("kassambara/ggcorrplot")

corr <- round(cor(Model), 1)

library(ggcorrplot)

# Plot
ggcorrplot(corr)


```




```{r echo=FALSE, Multivariate_Plots_three}


ggplot(subdf, aes(x = AvailableBankcardCredit, y = MonthlyLoanPayment, 
           color = as.character(LoanStatusNum))) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Loan Status',
                                          override.aes = list(alpha = 1,
                                                              size = 2)))+
  ggtitle('The impact of availabelity of Bank card credit and monthly loan
          payment on Loan status') +
  scale_x_continuous(limits = c(0, 10000), 
                     breaks = c(0, 2000, 4000, 6000, 8000)) + 
  scale_y_continuous(limits = c(0, 1000), 
                     breaks = c(0, 200, 400, 600, 800))  


```

This graph does not indicate any obvious pattern.



```{r echo=FALSE, Multivariate_Plots_four}

ggplot(subdf, aes(x = AvailableBankcardCredit, y = MonthlyLoanPayment, 
           color = as.character(LoanStatusNum))) +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Loan Status',
                                          override.aes = list(alpha = 1,
                                                              size = 2)))+
  ggtitle('The impact of availabelity of Bank card credit and monthly loan
          payment on Loan status') +
  geom_smooth(aes(color = as.character(LoanStatusNum))) + 
  scale_x_continuous(limits = c(0, 10000), 
                     breaks = c(0, 2000, 4000, 6000, 8000)) + 
  scale_y_continuous(limits = c(0, 1000), 
                     breaks = c(0, 200, 400, 600, 800)) 



```

To some extent, better loan status are associated with lower monthly payment.
The available Bank card credit does not affect the loan status.



```{r echo=FALSE, Multivariate_Plots_five}

ggplot(aes(y = CreditGradeNum, x = LoanStatusNum, 
           fill = LoanOriginalAmount),  data = subdf) + 
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))

```

too many missing values.
However, it looks like credit grade increases with loan original amount.



```{r echo=FALSE, Multivariate_Plots_six}

ggplot(aes(x = LoanOriginalAmount, y = LenderYield, 
           color = as.character(CreditGradeNum)),
       data = subdf) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Loan Status', 
                                  override.aes = list(alpha = 1, size = 2)))+
  scale_x_continuous(limits = c(0, 26000)) +    
  scale_y_continuous(limits = c(0, 0.4)) + 
  ggtitle('Credit quality and Loan original ammount')


```

Loan original value is to some extent discrete. Most of the loans are taking 
the value of $5000 and its duplicates (10K, 15K, 20K and so on).

It looks like that, bad loans (with low loan status level) are generating 
higher lender yield. This could be due to high interest rate.

For good loans (paid on schedule), the lender yield is associated with loan
original value.




```{r echo=FALSE, Multivariate_Plots_seven}


m1 <- lm(LenderYield ~ I(AvailableBankcardCredit), data = subdf)
m2 <- update(m1, ~ . + LoanOriginalAmount)
m3 <- update(m2, ~ . + MonthlyLoanPayment)
m4 <- update(m3, ~ . + InquiriesLast6Months)
m5 <- update(m4, ~ . + OnTimeProsperPayments)
mtable(m1, m2, m3, m4, m5)


```

The model can only explain 26% of the variance.
I tested many combinations including, for example, StatedMonthlyIncome
however, it did not enhance the R^2 of the model.


#Testing a model

First selecting a random sample from the dataframe

```{r echo=FALSE, Multivariate_Plots_eight}
sampledf <- as.data.frame(sample_n(na.omit(Model), 10))
sampledf
colnames(sampledf)
```



```{r echo=FALSE, Multivariate_Plots_nine}
#Testing a model

for(i in sampledf){
   y1= sampledf$subdf.LenderYield
   sample_i = 
     data.frame(
       "AvailableBankcardCredit" = sampledf$subdf.AvailableBankcardCredit,
       "LoanOriginalAmount" = sampledf$subdf.LoanOriginalAmount , 
       "MonthlyLoanPayment" = sampledf$subdf.MonthlyLoanPayment , 
       "InquiriesLast6Months" = sampledf$subdf.InquiriesLast6Months,
       "OnTimeProsperPayments" = sampledf$subdf.OnTimeProsperPayments)
      modelestimatefiti <- as.data.frame(predict(m5, newdata = sample_i, 
                                              interval="prediction", 
                                              level = .95))
   i = ifelse(y1 <= modelestimatefiti$upr &  y1 >= modelestimatefiti$lwr, 
          1, 0)
   print(i)
   }


```

### Double check
The lender yield of the first random sample is: 0.19

```{r echo=FALSE, Multivariate_Plots_ten}
#Testing a model

sample = data.frame(AvailableBankcardCredit = 3500, 
                    LoanOriginalAmount = 6000, 
                    MonthlyLoanPayment = 222.98, 
                    InquiriesLast6Months=4,
                    OnTimeProsperPayments = 14)

modelEstimate = predict(m5, newdata = sample, interval="prediction", 
                        level = .95)

#calculate the lender yield using the model
modelEstimate

```



# Multivariate Analysis

To some extent, better loan status is associated with lower monthly payment.
The available Bank card credit does not affect the loan status. However, it 
does affect the yield on loans.
It looks like credit grade increases with original loan amount. Loan original 
amount is to some extent discrete. Most of the loans are taking the value 
of $5000 and its duplicates (10K, 15K, 20K and so on).

Bad loans (with low loan status level) are generating higher lender yield. 
This could be due to penalties (for example) or higher interest rates.  
For good loans (paid on schedule), the lender yield is associated with original
loan amount.

A model for predicting the yield on loans was generated using five variables. 
The developed model can explain 26% of the variance. I tested many 
combinations including, for example, StatedMonthlyIncome. However, it did not
enhance the R^2 of the model.

The model was tested using randomly selected ten samples. The model perfectly
predicts the lender yield within the upper and the lower 95% confidence 
intervals.

Indeed, more investigation is needed to intensively evaluate the accuracy of
the model and assess other model parameters.





=================================================================


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}

p1=
qplot(data= subdf, y= LoanOriginalAmount, xlab = 'Loan original amount ($)',
      geom = 'boxplot', ylab ="") +
  ylim(0, quantile(subdf$LoanOriginalAmount, 0.99))

p2=
qplot(data= subdf, y= StatedMonthlyIncome, xlab = 'Stated Monthly income ($)',
      geom = 'boxplot', ylab= "") +
  ylim(0, quantile(subdf$StatedMonthlyIncome, 0.99))



grid.arrange(p1, p2, ncol = 2, top = textGrob("Loan and borrowers main characterestics",gp=gpar(fontsize=20,font=3)))

```

### Description One
Most of the borrowers are employees. The majority of the customers have low 
dept to income ratio (75% are below 0.32) and low Available Bank Card Credit 
(75% below $13,245). 
This is expected because the maximum loan in the data has the value of $35,000.
With such small loans it is not expected that the majority of the customers 
are coorporate entities.





### Plot Two
```{r echo=FALSE, Plot_Two}
p1 = 
ggplot(subdf, aes(x = AvailableBankcardCredit, y = MonthlyLoanPayment, 
           color = as.character(LoanStatusNum))) +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Loan Status',
                                          override.aes = list(alpha = 1,
                                                              size = 2)))+
  geom_smooth(aes(color = as.character(LoanStatusNum))) + 
  scale_x_continuous(limits = c(0, 10000), 
                     breaks = c(0, 2000, 4000, 6000, 8000)) + 
  scale_y_continuous(limits = c(0, 1000), 
                     breaks = c(0, 200, 400, 600, 800))+
  xlab("Available bank card credit ($)")+
  ylab("Monthly loan payment ($)")

p2 = 
ggplot(aes(x = LoanOriginalAmount, y = LenderYield, 
           color = as.character(CreditGradeNum)),
       data = subdf) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Loan Status', 
                                  override.aes = list(alpha = 1, size = 2)))+
  scale_x_continuous(limits = c(0, 26000)) +    
  scale_y_continuous(limits = c(0, 0.4)) + 
  xlab("Loan original amount ($)")+
  ylab("Lender yield")


grid.arrange(p1, p2, ncol = 2, 
             top = textGrob("Variables that impact loan status and lender yield",gp=gpar(fontsize=20,font=3)))



```

### Description Two

To some extent, better loan status are associated with lower monthly payment.
Loan original value is to some extent discrete. Most of the loans are taking 
the value of $5000 and its duplicates (10K, 15K, 20K and so on).

It looks like that, bad loans (with low loan status level) are generating 
higher lender yield. This could be due to high interest rate.

For good loans (paid on schedule), the lender yield is associated with loan
original value.


### Plot Three
```{r echo=FALSE, Plot_Three}
library(grid)

p1= 
  ggplot(subdf, aes(y = LenderYield, x = AvailableBankcardCredit)) + 
  geom_point(alpha = 1/50) +
  geom_smooth(method="lm")+
  ylab("Lender yield (%)")+
  xlab("Available bank card credit ($)")


p2= 
  ggplot(subdf, aes(y = LenderYield, x = LoanOriginalAmount)) + 
  geom_point(alpha = 1/50) + 
  geom_smooth(method="lm")+
  ylab("Lender yield (%)")+
  xlab("Loan original amount ($)")


p3=   
  ggplot(subdf, aes(y = LenderYield, x = MonthlyLoanPayment)) +
  geom_point(alpha = 1/50) + 
  xlim(0, 2000) + 
  geom_smooth(method="lm")+
  ylab("Lender yield (%)")+
  xlab("Monthly loan payment ($)")


p4 = 
  ggplot(subdf, aes(y = LenderYield, x = InquiriesLast6Months)) +
  geom_point(alpha = 1/50) + 
  xlim(0, 15) + 
  geom_smooth(method="lm")+
  ylab("Lender yield (%)")+
  xlab("Inquiries in the last 6 months")



p5 = 
  ggplot(subdf, aes(y = LenderYield, x = OnTimeProsperPayments)) +
  geom_point(alpha = 1/20) + 
  xlim(0, 80) +
  geom_smooth(method="lm")+
  ylab("Lender yield (%)")+
  xlab("On time prosper payments")



grid.arrange(p1, p2, p3, p4, p5, ncol = 2, 
             top = textGrob("Variables used in the model",
                            gp=gpar(fontsize=20,font=3)))



```

### Description Three

Lender yield is increasing with Inquiries last 6 months and ontime prosper 
payment. 
Lender yield is decreasing with loan original amount, Employment status 
duration, current credit lines and stated monthly income.

------

# Reflection

The developed model has an R square of 26% which is quite fair. However, when 
the model was tested using 10 random samples from the data, it was successful 
in predicting lender yield within the 95% confidence interval. 
This is quite impressing.

I limited the time to invest in this project to 30 working hours. If I had more
time, I would continue in developing other models to predict other variables of
interest.

There are many limitations for this model. First, the maximum loan value was
only 35K. This model might work well with small personal loans. For corporate 
and large loans, more investigations must be conducted before reaching 
premature conclusions. Another limitation, I focussed only on a linear 
relationship.
Probably exponential relationships exist in the data and revealing them might 
nhance the model accuracy.




# References

- Source of the Data
https://s3.amazonaws.com/video.udacity-data.com/topher/2018/September/5babed58_data-set-options-explore-and-summarize-data-project/data-set-options-explore-and-summarize-data-project.pdf

- Variable definitions
https://rpubs.com/mohohoh/ProsperDefinitions

- Code for random subset of a dataframe
https://www.statmethods.net/management/subset.html

- R ifelse() Function
https://www.datamentor.io/r-programming/ifelse-function/

- Sources for pie chart code
https://stackoverflow.com/questions/33594642/beautiful-pie-charts-with-r

http://www.sthda.com/english/wiki/ggplot2-pie-chart-quick-start-guide-r-software-and-data-visualization

- Source for the code of the correlation matrix
http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2


- Source for teh forcats code and teh bar plot
https://stackoverflow.com/questions/42497063/r-sorting-factor-by-level-frequency-and-plotting

https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2


